The evolution of gene expression in the term placenta of viviparous mammals

### 2015-11-20 manuscript/manuscript.Rnw
work on GSE analyses

% SDC4 -- proteoglycan

@@

\subsection{topGO Analyses of Individual Transcriptomes}

In order to identify major functional categories of genes expressed in
the individual transcriptomes, we undertook functional annotation
clustering analyses using topGO. We examined \ac{GSE} in 1:1 human
orthologs which had \ac{FPKM} values in the top 5\% of expressed genes
%
Significant enrichment in genes ($\mathrm{FDR}\le0.05$) in the top 5\%
of \ac{FPKM} values in at least 12 of the
\Sexpr{length(species.names)} species included clusters such as
membrane organization , locomotion, tissue development, extracellular
matrix organization, regulation of peptidase activity, and various
cell death related pathways among others
(\cref{tab:go_terms_all_species}).

(Supplemental %
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}.) %

Human shows enrichment in clusters such as collagen-binding, hormone
activity, female pregnancy, coagulation, and the defense response.
Bonobo shows enrichment in clusters related to apoptosis,
cytoskeleton-related genes, and the stress response. Cow shows only
four significantly enriched clusters; however many genes highly
expressed in the cow are ruminant-specific placenta genes and are not
yet annotated to specific GO categories. Dog shows enrichment in genes
associated with RNA, as well as homeostasis, glycoproteins, lyosome,
and calcium binding. Elephant shows enrichment in genes associated
with RNA, coagulation and calcium binding. Opossum shows enrichment in
genes involved in apoptosis, oxidative stress, mitochondrion, hormone
metabolism, and lipids.

>>>>

% SDC4 -- proteoglycan (syndecan) cell adhesion

% LAMC1 -- laminin

% LUM -- proteoglycan

% VDAC1 -- mitochondrial gene?

% F11R -- tight junctions

@@

\subsection{\ac{GSE} Analyses of Individual Transcriptomes}

In order to identify major functional categories of genes expressed in
the individual transcriptomes, we examined \ac{GSE} in 1:1 human
orthologs which had \ac{FPKM} values in the top 5\% of expressed genes
(Supplemental %
\cref{tab:go_terms_enriched_top_five_dasypus_novemcinctus,tab:go_terms_enriched_top_five_equus_caballus,tab:go_terms_enriched_top_five_homo_sapiens,tab:go_terms_enriched_top_five_loxodonta_africana,tab:go_terms_enriched_top_five_monodelphis_domestica,tab:go_terms_enriched_top_five_mus_musculus,tab:go_terms_enriched_top_five_ovis_aries,tab:go_terms_enriched_top_five_pan_paniscus,tab:go_terms_enriched_top_five_sus_scrofa,tab:go_terms_enriched_top_five_ateles_fusciceps,tab:go_terms_enriched_top_five_nannospalax_galili,tab:go_terms_enriched_top_five_spalax_carmeli,tab:go_terms_enriched_top_five_bos_taurus,tab:go_terms_enriched_top_five_canis_familiaris}) using topGO \cite{topGO:2.20.0}. %
%
Significant enrichment in genes ($\mathrm{FDR}\le0.05$) in the top 5\%
of \ac{FPKM} values in at least 12 of the
\Sexpr{length(species.names)} species included clusters such as
membrane organization , locomotion, tissue development, extracellular
matrix organization, regulation of peptidase activity, and various
cell death related pathways among others
(\cref{tab:go_terms_all_species}). If we instead examine the top 100
genes by \ac{FPKM} in each of the \Sexpr{length(species.names)}
species (Supplemental %
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris})
notable enrichments include extracellular matrix, female pregnancy,
integrin and collagen-related processes, and various cell death
related processes.

% \textit{Ateles fuciceps} shows enrichment in extracellular
% matrix and structural molecular activity (Supplemental
% \cref{tab:go_terms_enriched_ateles_fusciceps}). \textit{Homo sapiens}
% placentas are enriched in female pregnancy, estrogen biosynthesis,
% reproductive process, hormone aactivity, oxygen transporter activity,
% and haptoglobin binding (Supplemental
% \cref{tab:go_terms_enriched_homo_sapiens}). \textit{Pan paniscus}
% placentas are enriched in clusters related to stress, apoptosis, and
% response to abiotic stimuli. \textit{Mus musculus} placentas are
% enriched in cell junction, vesicle transport, in utero embryonic
% development, and many cytoskeletal processes. \textit{Nannospalax
%   galili} placentas are enriched in extracellular matrix organization,
% integrin binding, and cell death.
% 
% Cow shows only four
% significantly enriched clusters; however many genes highly expressed
% in the cow are ruminant-specific placenta genes and are not yet
% annotated to specific GO categories. Dog shows enrichment in genes
% associated with RNA, as well as homeostasis, glycoproteins, lyosome,
% and calcium binding. Elephant shows enrichment in genes associated
% with RNA, coagulation and calcium binding. Opossum shows enrichment in
% genes involved in apoptosis, oxidative stress, mitochondrion, hormone
% metabolism, and lipids.
